Approximation as Differentiation

### 2025-05-24 main/introduction.tex
Minor reorg of intro.

Provenance tracking aims to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Program slicing is a collection of techniques for provenance tracking that attempts to take a run of a program and and areas of interest in the output and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

Provenance tracking is important because in order to audit any computational process, we need robust and well-founded notions of provenance to track how data are used, how data are computed, and what data influenced a particular outcome. Existing approaches are often tied to particular programming languages or implementations. In this paper we develop a general categorical approach.

\note{We will use biproducts, and use some results from \cite{karvonen2020}.}

@@

Perera and collaborators introduced the idea of {\em Galois Program Slicing} as a particular conception of program slicing for provenance, described in several publications~\cite{perera12a,perera16d,ricciotti17}. Galois program slicing forms the basis of Perera's data visualisation tool \href{https://f.luid.org/}{Fluid} that allows interactive exploration of programmatically generated visualisations.

At a high level, Galois Program Slicing assumes that, for each possible value that may be input or output by a program, there exists a lattice of {\em approximations} of that value. For a particular run of a program that takes input $x$ and produces output $y$, we also get a Galois connection between the lattice of approximations of $x$ and the lattice of approximations of $y$. The right half of the Galois connection is the ``forward direction'' taking approximations of the input to approximations of the output; the left half of the Galois connection is the ``backward direction'' that takes approximations of the output to the least (i.e., most approximate) approximation of the input that gives rise to this output approximation.

>>>>

Provenance tracking aims to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Provenance tracking has a wide range of applications, from debugging and program comprehension to reproducibility and transparency in scientific workflows. In order to audit any computational process, we need robust and well-founded notions of provenance to track how data are used, how data are computed, and what data influenced a particular outcome. \emph{Program slicing} is a collection of techniques for provenance tracking that attempts to take a run of a program and areas of interest in the output, and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

Existing approaches to program slicing are often tied to particular programming languages or implementations. In this paper we develop a general categorical approach to program slicing and \todo{...}.

@@

Perera and collaborators introduced the idea of {\em Galois Program Slicing} as a particular conception of program slicing for provenance, described in several publications~\cite{perera12a,perera16d,ricciotti17}. Galois Program Slicing (hereafter simply {\emph \GPS}) forms the basis of the open source data visualisation tool \href{https://f.luid.org/}{Fluid} that allows interactive exploration of programmatically generated visualisations.

At a high level, \GPS assumes that, for each possible value that may be input or output by a program, there exists a lattice of {\em approximations} of that value. For a particular run of a program that takes input $x$ and produces output $y$, we also get a Galois connection between the lattice of approximations of $x$ and the lattice of approximations of $y$. The right half of the Galois connection is the ``forward direction'' taking approximations of the input to approximations of the output; the left half of the Galois connection is the ``backward direction'' that takes approximations of the output to the least (i.e., most approximate) approximation of the input that gives rise to this output approximation.
